---
title: "Logicals and Functions"
subtitle: "R Syntax 2"
output: html_notebook
---



## Functions

Most things you can do with R work with named functions, rather than symbol operators.  For example, there is no symbol to calculate the square root; instead, you would use the `sqrt()` function like so: `sqrt(4)`.  All functions in R have the following syntax:

`output <- functionName(input)`

Let's try out some math functions to get a feel for the syntax!

Math functions:
  - Positive/Negative functions: `abs()`, `sign()`, 
  - Rounding functions: `ceiling()`, `floor()`, `round()`
  - Power Transforms: `sqrt()`, `exp()`, `log()`, `log10()`, `log2()`, `log1p()`
  - Trigonometry functions: `sin()`, `cos()`, `tan()`, `asin()`, `acos()`, etc.
  

**Tip**: Not sure what a function does?  Get the help text by putting a ? before it:
`?abs`

**Tip2**: R not responding?  If you see a + sign at the beginning of the console,
it means R doesn't think the command is finished.  Cancel the command by clicking
on the cancel and press the "Esc" key.  When you see a > sign, R is ready for new
commands.


**Exercises**

Convert the following sentences into R code.  Each sentence may involve multiple
calculations, and there may be multiple ways of getting the same answer.


Set "x" to "The absolute value of negative four"
```{r}
x <- abs(-4)
```


Set "result" to "4.2 rounded to the nearest integer is less than 4.5"
```{r}

```


Set "data" to "4.2 rounded up is less than 4.5"
```{r}
round(digits=2, x=4.1234566)
```


Set "ninety" to "the sine of pi over two is equal to one"
```{r}

```


Set "a" to "the natural log of thirty is greater or equal to 10"
```{r}

```

Set "b" to "ten to the power of the log-base-10 of eighty is equal to eighty"
```{r}
format(10 ^ log10(80), nsmall = 20)
```

```{r}
format(.1 + .2, nsmall = 20)
```

Set "c" to "the square root of the sum of the square of three and the square of four
is equal to the square root of twenty-five"
```{r}

```

Set "vera" equal to "e to the power of five"  (Tip: `exp()`)
```{r}

```


## Making New Functions

Sometimes you need to do the same set of calculations over an over again.  
Rather than writing out the whole code each time, you can summarise those calculations by defining your own function and using that.  Here is how you
do that in R (note the curly braces):

```
functionName <- function(input) {
    output <- input + 3
    return(output)
}

functionName(4)
functionName(5)
```

Note: the `return()` part is optional.  R will automatically return the output
of the last calculation (just like how the last calculation is automatically
printed in the notebook).  So you could write the same thing like this:

```
functionName <- function(input) {
    output <- input + 3
    output
}

functionName(4)
functionName(5)
```

Or on a single line:


```
functionName <- function(input) {input + 3}
functionName(4)
functionName(5)
```


**Exercises**

Let's get some practice making functions!  Please make the requested functions 
that pass the tests below.  If the `stopifnot()` lines are false, you will
see an error message.  If they are true, `stopifnot()` won't return anything,
and you will see the "success" message printed.


Make the function `isThree()` that returns True if the input is three, False if not.
```{r}

isThree <- function(x) {x == 4}
stopifnot(
  isThree(3) == TRUE,
  isThree(4) == FALSE
)
print("Success!")
```


Make the function `isFour()` that returns TRUE if the input is four, False if not.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  isFour(3) == FALSE,
  isFour(4) == TRUE
)
print("Success!")
```


Make the function `isPositive()` that returns TRUE if the input is greater or equal to zero, False if not.  (in this case, we'll pretend zero is positive)

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  isPositive(3) == TRUE,
  isPositive(-2) == FALSE,
  isPositive(0) == TRUE
)
print("Success!")
```


Make the function `square()` that returns the square of the input.
```{r}

# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  square(3) == 9,
  square(4) == 16
)
print("Success!")
```

Make a function `not()` that returns False if the input is True, True if False

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  not(T) == F,
  not(F) == T
)
print("Success!")
```


### Functions with multiple inputs

Functions can have as many inputs as you want!  For example:

```
add <- function(x, y) { x + y }
add(2, 5)
add(4, 6)
```

**Exercises**


Make a function `subtract()` that subtracts the second input from the first.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  subtract(5, 3) == 2,
  subtract(2, 5) == -3,
  subtract(10, 5) == 5
)
print("Success!")
```


Make a function `isGreater()` that returns True if the first input is greater than the second, False if not.

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  isGreater(3, 4) == FALSE,
  isGreater(4, 3) == TRUE,
  isGreater(4, 4) == FALSE
)
print("Success!")
```


Make a function `sumOfSquares()` that adds together the squares of each input.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  sumOfSquares(1, 2) == 5,
  sumOfSquares(2, 3) == 13,
  sumOfSquares(3, 4) == 25
)
print("Success!")
```


Make a function `average()` that calculates the mean of the three inputs (the sum of the inputs, divided by the number of inputs)
```{r}
# Function here:
source("../cookbook.R")

# Run the Tests below to see if the function works as requested
stopifnot(
  theMean(3, 3, 3) == 3,
  theMean(3, 4, 5) == 4,
  theMean(5, 0, 10) == 5,
  theMean(5, 3) == 3
)
print("Success!")
```




## Functions with default input values

Often, functions will have some inputs "pre-set" to a certain value.
This makes the input "optional"; you don't have to give it if you are happy
with the default.  Any inputs that don't have default values will still be 
required for the function to run.

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
add <- function(x, y=1){x + y}
add(3)  # 4
add(4)  # 5
add(3, 4) # 7
add(3, y=4)  # 7
add(x=3, y=4)  # 7
add(y=4, x=3)  # 7
```


**Exercises**


Make the function `addN()` that adds 3 to whatever number is supplied by default,
but can also add other numbers.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
addN <- function(x, y=3){x + y}

stopifnot(
  addN(4, 3) == 7,
  addN(5) == 8,
  addN(6, y=1) == 7,
  addN(7, 8) == 15
)
print("Success!")
```



Make the function `subN()` that subtracts zero from whatever number is supplied
by default, but can also subtract other numbers.

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  subN(4, 3) == 1,
  subN(5) == 5,
  subN(5, y=1) == 4,
  subN(7, 8) == -1
)
print("Success!")
```



Most functions with defaults use them as settings that change the way the function
works.  Usually, these defaults are set to TRUE or FALSE, and you can change them
to the opposite.  To do that, we'll need to use if-else statements inside the function.

Single-line example:
```
isPositive <- function(x, incZero=F){x > 0 if incZero == F else x >= 0}
isPositive(0)  # FALSE
isPositive(0, incZero=T)  # TRUE
```

Multi-line example:
```
isPositive <- function(x, incZero=F){
  if (incZero == FALSE) {
    result <- x > 0
  } else {
    result <- x >= 0
  }
  result
}
```

Another way is with the `ifelse()` function. Here's an example:
```
isPositive <- function(x, incZero=F){ifelse(incZero == F, x > 0, x >= 0)}
```


Make the function `sub2()` that subtracts the first number minus the second 
normally, but subtracts the second number minus the first if "reverse" is True 
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  sub2(3, 2) == 1,
  sub2(2, 3) == -1,
  sub2(2, 3, reverse=T) == 1,
  sub2(1, 4, reverse=T) == 3
)
print("Success!")
```


Make the function `math()` that adds two numbers if "add" is True and subtracts
them if "add" is False

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  math(2, 3) == 5,
  math(5, 3, add=T) == 8,
  math(4, 5, add=F) == -1,
  math(1, 1, add=F) == 0
)
```


Make the function `roundnum()` that rounds a number up if "ceil" is True and
rounds it to the nearest integer if "ceil" if False
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  roundnum(5.1) == 5,
  roundnum(5.9) == 6,
  roundnum(5.1, ceil=T) == 6,
  roundnum(4.2, ceil=T) == 5
)
```




## Extra (Optional): Making Functions with a variable number of inputs

Many functions in R have a special sumbol as an input: the "Elipsis" `...`

The ... is a special variable that holds any extra positional inputs. It can
be used to create functions that accept unlimited inputs, which the function
often makes a vector out of internally for calculation.

For example, one can make a function that accepts a "data" vector:
```{r}
average <- function(data){
  mean(data)
}

average(c(1, 2, 3))  # Making the vector when the function is called.
```

Alternatively, one can make a function that builds a "data" vector internally:
```{r}
average <- function(...){
  data <- c(...)  # Making a vector inside the function.
  mean(data)
}

average(1, 2, 3)
```

You can combine the ... symbol with positional inputs and optional inputs.
If R is not sure where to put the input, it will stick it in the ...
```{r}
showInputs <- function(x, ..., y=100){
  print(x)
  print(c(...))
  print(y)
}

showInputs(1, 2, 3, 4, 5)
```


**Exercises**

Make an `add()` function that takes a variable number of inputs to add.
```{r}
# function here

# tests here
stopifnot(
  add(1) == 1,
  add(1, 2) == 2,
  add(2, 3) == 5,
  add(1, 2, 3) == 6,
  add(1, 2, 3, 4) == 10
)
print("Success!")
```




Make a `math()` function that adds all the inputs if "add" is True, and multiplies them if not.
```{r}
# function here

# tests here
stopifnot(
  math(1, 2) == 3,
  math(1, 2, add=F) == 2,
  math(1, 2, 3, 4, add=T) == 10,
  math(1, 2, 3, 4, add=F) == 24
)
print("Success!")

```



Make an `isPositive()` function that returns a vector with a value for True for each positive number, and False for each nonpositive number.
If "incZero" is True, then 0 is considered positive.  (Tip: Don't use the `ifelse()` function.)
```{r}
# function here


# tests here
stopifnot(
  identical(isPositive(1, 2), c(T, T)),
  identical(isPositive(-1, 0, 1), c(F, F, T)),
  identical(isPositive(-1, 0, 1, incZero=T), c(F, T, T))
)
print("Success!")
```

