---
title: "Vectors and Descriptive Statistics"
subtitle: "R Syntax 3"
output: html_notebook
---

## Vectors: Collecting Values Together into a Single Variable Name

In R, it is rare for you to create dozens of variables; usually, putting
data together as a dataset into a single variable makes work simpler!

Data is grouped together into a data structure.  R supports many kinds of data structures (e.g. matrices, lists, dataframes, etc), but the simplest one is
the *vector*, a simple sequence of values.  We can get started on many statistics
just using vectors, and most operations on vectors work on the other data 
structures, making vectors a great place to get started in R.


## Creating Vectors

Many functions exist in R to quickly create vectors:

  - **`c(1, 3, 5)`**: combine values into the vector  *(1, 3, 5)*.
  - **`1:5`**: create a vector of the values 1 through 5 *(1, 2, 3, 4, 5)*
  - **`seq(1, 5)`**: Also creates a vector of 1 through 5 *(1, 2, 3, 4, 5)*
  - **`seq(1, 5, by=2)`**:  1 through 5, every 2 values  *(1, 3, 5)*
  - **`seq(1, 5, length.out=2)`** Two total values from 1 through 5 *(1, 5)*
  

**Exercises**

Make a vector "data" with the values 2, 2, 5, 7, 2
```{r}
data <- c(2, 2, 5, 7, 2)
data
```

Make a vector "data" with the values 4, 6, 7, 3, 4
```{r}
x <- 1
y <- 2
```


Make a vector "data" with the values 5, 3, 2
```{r}

```


Make a vector "data" with the values T, F, F, T
```{r}

```


Make a vector "data" with the values 2.6, 3.2, 1.1, 0.9
```{r}

```



Without using the c() function, make a vector "data" with the values 1, 2, 3, 4
```{r}

```


Without using the c() function, make a vector "data" with the values 1, 3, 5, 7
```{r}

```


Without using the c() function, make a vector "data" with the values -20 through 20
```{r}

```

Make a vector "data" with 10 values evenly space from 1 to 35
```{r}

```


Make a vector with the values (1, 2, 3, 4.4, 5).  What is strange about this
vector?
```{r}

```

Makea vector with the values (T, F, 3, F, T).  What is strange about this vector?
```{r}

```


## Summary Statistics on Vectors

Many functions take sequences of data like vectors as an input.  They are really
useful, because they don't care how much data you have!

Let's start with these functions: `sort()`, `length()` `sum()`, `mean()`.


**Exercises**

Sort this dataset in ascending order (i.e. from low to high): (6, 2, 5, 3, 4)
```{r}

```

Using an optional input to the `sort()` function:
Sort this dataset in descending order (i.e. from high to low): (6, 2, 5, 3, 4)
```{r}

```



Calculate the length of the vector (1, 4, 5).  What does the length function do?
```{r}

```

Calculate the sum of all the values from -5 to 5.
```{r}

```

Calculate the sum of all the values from one to one million!  Does it take long?
```{r}

```

Calculate the mean of the dataset (1, 5, 3, 2).
```{r}

```


Without using the `mean()` function, calculate the mean of the dataset (1, 5, 3, 2)
*Tip*: The mean is the sum of all the values divided by the total number of values.
```{r}

```


## Vectorized Tranformations on Vectors: R's Cycle Rule

All of the math and logical operations in R (in fact, most operations in R)
work as "vectorized" operations.  This means that if a calculation is meant for
one value, it is automatically repeated for every value in the data structure.

For example, to add 1 to all values from 10 to 15 is as simple as `10:15 + 1`
Let's try it out!


**Exercises**

Multiply all the values from 1 to 5 by 10.
```{r}
1:5 * 10
```


Subtract 1000 from all the values from 1 to 10.
```{r}

```


Check each value from 1 to 10 if it is greater than 5 (tip: answer should have 5 Fs and 5 Ts)
```{r}

```


Check each value from 1 to 10 if the value's square (power of 2) is less than 10.
```{r}

```


calculate the square root of all the values from 1 to 10.
```{r}

```


Calculate the sum of squares of the values 1 through 10

```{r}

```


Calculate the square root of the sum of squares of the values 1 through 10

```{r}

```



When math operations like the `sum()` or `mean()` are done on logical values (T and F),
R automatically converts them to 1 and 0, respectively.  This is a useful feature,
and helps to calculate **counts** and **proportions**.  Let's try it out!

How many values are there greater than 5 in this dataset: (1, 6, 3, 5, 7)?
```{r}
data <- c(1, 6, 3, 5, 7)
sum(data > 5)
```

What proportion (i.e. percentage) of values in the dataset are greater than 5?
(1 means 100%, and 0 is 0%)
```{r}
data <- c(1, 6, 3, 5, 7)
mean(data > 5)
```

How many values in the dataset (5, 7, -2, -4, -3) are negative?
```{r}

```


What proportion of values in the dataset 1 through 50 are greater than the
dataset's mean?
```{r}

```


How many values in the dataset (5, 8, 4, 6, 5, 7, 5) are equal to 5?
```{r}

```


What is the average height of the men in this dataset?  Of the women?
heights: (140, 150, 144, 128, 152, 148, 175, 164, 156, 149, 188, 201)
isMale:  (T, F, F, F, T, T, F, F, T, T, F, T)
```{r}

```


What proportion of the nucleotides in this DNA sequence is Cytosine?
*Note*: To make "character" (i.e. text, string) values, put single-quotes or double-quotes around the value
```{r}
dna <- c("C", "G", "A", "T", "G", "C", "T", "G")
dna
```


What proportion of the nucleotides in this DNA sequence is either Argenine?
*Note*: The `strsplit()` function splits up a single string into a vector of strings. 
```{r}
dna <- strsplit("CGTATATAGGGCAGGAT", "")[[1]]
mean(dna == "A") + mean(dna == "T")
mean(dna == "A" | dna == "T")



```




What proportion of the values from 1 to 100 are divisible by 2 
(i.e. has a remainder of 0 after dividing the value by 2)?
(reminder: the `%%` operator is used to get the remainder from integer division).
```{r}

```


## Cycle Rule: Vectorizing along multiple vectors

This vectorization rule also applies when working with multiple vectors, letting
you add up vectors together.  The way R does this is called the **"Cycle Rule"**;
it cycles through each value in the vectors, doing the operation on each in turn.
It continues cycling through values until the *longest* vector is finished.


Add up all the values in dataset "data1" with the values 1 through 5
and all the values in dataset "data2" with the values 21 through 25
```{r}
1:5 + 21:25
```

Same as above, but creating variables first.
```{r}
data1 <- 1:5
data2 <- 21:25
data1 + data2
```


Add up all the values between the values 1 through 5 and 10 through 14.
```{r}

```


Multiply these two datasets together: (-1, 0, 1, 0, -1) and (1, 2, 3, 4, 5)
```{r}

```



## Subsetting Data by Index: Selecting Values from Vectors

Once you've collected data together, you're not stuck with all of your data;
a "subset" (i.e. smaller part) of a dataset can also be extracted out!

There are several different ways to subset your data in R; let's start with
the most general one: the `[]` square brackets.


  - `data[N]` to get the Nth value  (for example, `data[3]`)
  - `data[indices]` to get multiple values at once  (for example, `data[c(1, 3, 5)]`)
  - `data[-N]` to get all the values *except* the Nth   (for example, `data[-3]`)
  
  - `data[N:P]`  to get all the values from the Nth to the Pth:   (for example, `data[3:6]`)
  - `data[N:length(data)]`  to get all the values from the Nth to the end  (for example, `data[3:length(data)]`)
  - `data[-N:-P]`  to get all the values *except* the Nth through Pth:   (for example, `data[-3:-6]`)
  
  
**Exercises**

Subset out the requested values from the `data` vector given below:
```{r}
data <- 11:30
data
```

Get the third value.
```{r}
data[3]
```

Get the fifth value.
```{r}

```


Get all the values except the second values.
```{r}

```

Get the first through fifth values.
```{r}

```


Get the tenth through fourteenth values.
```{r}

```


Get the tenth value to the end of the vector (note: have R calculate what the last index is for you)
```{r}

```


Get the fifth value through the first value (result should be 15, 14, 13, 12, 11)
```{r}

```


Reverse the order of the values
```{r}

```


Get the fourth, sixth, ninth, and tenth values.
```{r}

```


Get the second, tenth, and eleventh values.
```{r}

```

Get every second value, starting with the first value.  
(Hint: use the `seq()` function to make the vector of indices)
```{r}

```



## Subsetting Data with Logical Values


Unfortunately, most data doesn't come organized in a nice, orderly fashion, and 
so we can't know when writing the code that we want the Nth value using the 
techniques above.

Instead, we have to *search* for the values we want. R has a lot of tools for doing
this, and all of them involve three steps:


```{r}
# 1. Define the dataset
data <- c(2, 4, 3, 5)  # 
data

# 2. For each value in the dataset, determine if the value meets your search criteria.
# (Turns each value into T or F, based on if it is greater than 3)
isBig <- data > 3 
isBig


# 3. Subset out only the True values into a new vector.
bigData <- data[isBig]  # S
bigData
```

You can shorten the code by combining the last two steps (two ways shown below):
```{r}
# Approach 1: []
data[data > 3]

# Approach 2: the subset() function
subset(data, data > 3)
```




**Exercises**

Using the data provided below and the `length()`, `sum()`, and`mean()` functions,
in the following list, please get a subset of the data and answer the following questions:


Dataset:
```{r}
data <- c(1, 4, 8, 4, 9, 3, 2, 8, 3, 8, 32, 8, 2, 7, 3, 7, 1, 32, 4)
data
```

Which values in the dataset are greater than 8?  How many are there?
```{r}
goodData <- data[data > 8]
goodData
length(goodData)
```

Which values in the dataset are less than 7?  How many are there?
```{r}

```


Which values in the dataset are between 3 and 6?  What is the mean of their values?
```{r}

```


What proportion of the dataset is equal to 8?
```{r}

```


What proportion of the dataset is between 4 and 9?
```{r}

```


What proportion of the first 10 values in the dataset is less than 5?
```{r}

```


What proportion of the last 10 values in the dataset is less than 5?
```{r}

```




