---
title: "One-Dimensional Sequences: Named Vectors"
output: html_notebook
---

# Vectors: Positional Data


# Named Vectors

You can keep the data values and labels together!  Instead of having 
positional vectors, you can give each index a name:

```{r}

```

Subsetting works by name.  
```{r}

```

If you want only the number from the subset, you can use the double square brackets:
```{r}

```


Plots will automatically use the names:
```{r}
barplot(heights)
```


Finally, you can separate out the names from the values with the `names()` and `as.numeric()` functions:
```{r}
names(heights)
as.numeric(heights)
```


**Exercises**
Let's try it out!  Make a named "age" vector for the following children:
  - Cindy: Age 5
  - Anna: Age 4
  - Emma: Age 6
  - Adrian: Age 8

```{r}

```


Use R to find the values:  What is Emma's age?
```{r}

```


Use R to find the values:  What is Anna's age?
```{r}

```


Use R to find the values:  What is the `mean()` age of Cindy and Emma?
```{r}

```


Use R to find the values:  What is the `mean()` age of Adrian and Anna?
```{r}

```


Make a bar plot of the children's ages, including title, axis labels, and labels for each bar.
```{r}

```


Remove Anna from the dataset, and make a new bar plot.
```{r}

```
  

