

# Nonparametric Statistics: Calculating Proportions and Percentages

Dataset: The number of bacteria in a sample of water samples
```{r}
bacteria_count <- c(161, 136, 109, 124, 164, 123, 147, 135, 139, 169, 148, 115, 155, 110)
```


### A
What proportion of the water samples had less than 120 bacteria?
```{r}

```

### B
What proportion of the water samples had less than 180 bacteria?
```{r}

```


### C
What percent of the water samples had less than 300 bacteria?
```{r}

```

### D
What is the value of the dataset that reaches the 100% mark? (i.e. what is the maximum of the data)?
```{r}

```


### E
What is the value of the dataset, just below which would contain 0% of the data? (i.e. what is the maximum of the data)?
```{r}

```


# Calculating proportions visually

## Estimating from a distribution of data points
### F
`plot()` Reading from the graph below, answer the following
questions:
  a) What proportion of the data does 147 bacteria samples lie at?
  b) What proportion of the data does 110 bacteria samples lie at?
  c) How many samples below which does 50% of the bacteria samples lie at?
```{r}
# plot(bacteria_count)
```

How easy was this to do?  How accurately did you feel you answered the question?

## Measuring from a sorted distribution of data points
### G
Let's try this again, but this time first `sort()`ing the data so that the lowest value
comes first and the highest value comes last:
  a) What proportion of the data does 147 bacteria samples lie at?
  b) What proportion of the data does 110 bacteria samples lie at?
  c) How many samples below which does 50% of the bacteria samples lie at?
```{r}

```


Simpler and more accurate, right?  A lot of *nonparametric* statistics is based around the numbers that come out of sorting data and looking at the order they come in.  Below,
we'll use these values to calculate percentages, percentiles, and generate
emperical cumulative distribution functions (ECDFs) and box plots.


# `sort()` and `order()` Calculating all the percentages!

### H
Use the `order()` function on the dataset to calculate the order of the values, for
each sample in the dataset.  Which value is the first-highest?  Which is the fifth-highest?
```{r}
rbind(
  bacteria_count, 
  order(bacteria_count),
  rank(bacteria_count),
  bacteria_count[order(bacteria_count)]
  
)

bacteria_count[order(bacteria_count) == 1]
bacteria_count[rank(bacteria_count) == 1]


```

### I
What is the order of the values if you `sort()` them first? What if you sort them
in descending order?
```{r}

```

### J
Convert the orders to percentages, by dividing them from the length of the dataset
(or the maximum order value):
```{r}

```


# Percentiles

The *"percentile"* is like the percentage, but its unit is the same as the dataset.
To understand the difference:
  - *"Percentage"*: "What percentage of the data is less than 150 bacteria?"
  - *"Percentile"*: "How many bacteria are at the 40% mark?"
  
We've probably thought about percentile most when talking about class performance:
  - *Percentage*: "I answered 90% of the questions correctly!"
  - *Percentile*: "My exam score was the same or better thn 90% of my class!"
  
Percentiles are very helpful for cutting up your data into equal-sized pieces,
called *"quantiles"*. If you want to divide your students into 2 quantiles, you could
find the 50th percentile value (a.k.a. the *"median"*), and put all the students into one group if their score is above that point, and the rest in the other group.  You'd
have (roughly) the same number of students in each group!
  - Want 3 quantiles?  Divide the data at the 33%, 66%, and 100% points.
  - Want 5 quantiles?  Divide the data at the 20%, 40%, 60%, 80%, and 100% points.
  
The main takeaway is that percentiles help you find the cutoff values so you can 
cut your dataset into fixed-sized sub-datasets.

## Estimating Percentiles Visually: data on x, percentages on y
First, let's do the visual estimation exercise again, but this time with the
values inversed, to show a more traditional graph (labels on the x-axis, calculation on the y-axis).  This will be the bases for the ECDFs we'll be making, and getting a feel for
interpreting them will be helpful.d

### K
`plot()` the sorted bacteria sample values, so that the x-axis are the bacteria
values and the y-axis is the percentage of the data they lie at.  Just looking
at this graph, answer the following questions:
  a) What proportion of the data does 140 bacteria samples lie at?
  b) What proportion of the data does 200 bacteria samples lie at?
  c) How many samples below which does 50% of the bacteria samples lie at?
```{r}

```

## Using R to calculate percentiles

R's `quantile()` function makes it easier to calculate percentiles.  Let's try it out!

### L
Example: What value would lie at the 30th percentile mark?
```{r}
quantile(bacteria_count, .3)
```

### M
What value would lie at the 50th percentile mark?  Is this the same value that the
`median()` function gives you?
```{r}

```

### N
Calculate multiple percentilces at once by providing multiple probability values
to the quantile function.  Let's find the quantiles at the 20%, 50%, and 80% marks:
```{r}

```

### O
Use the `ecdf()` function to calculate the ECDF.  This will make a special object 
that when given to the `plot()` function wil make the ECDF plot.  Try it out here:
```{r}

```

### P
Use the `quantile()` function to make the ECDF plot yourself, generating a point for every
percentile value (i.e. 1%, 2%, 3%, ... 99%, 100%) and plotting it.  the data should look the same as what the `ecdf()` function creates.
```{r}

```


# Using Percentiles to Model Data

From this activity, we have found a tool that we can use to describe any dataset
statistically:
  - A Central Measure (the "Best Guess" for a value):  the 50% point (a.k.a. the *"median"*)
  
  
What's needed still is a way of estimating the amount of variation in the dataset.  
Luckily, we can get this by also using percentiles, this time two percentiles!
We can, for example, say that the middle 50% of the data is between the 25 percentile
and the 75th percentile.  That doesn't tell us about the other half of the data,
of course, but maybe that's okay for our purposes.  Want to be more complete?
We could talk about all 100% of the data, and report the values between the 0th percentile
and 100th percentile.  Some common ranges used are:
  - middle 50% (25 - 75)
  - middle 95% (2.5 - 97.5)
  - middle 99% (0.5 - 99.5)
Some other odd numbers commonly used are:
  - middle 68% (16 - 84)
  - middle 99.7% (.15- 99.85)
  


## Reporting Percentiles
  
We can represent this spread in two different ways:
  a. *with two percentiles*: (i.e. "80% of the data falls between 13.7 and 54.9.")
  b. *with a median and difference (Inter-Quantile-Range, IQR)* (i.e. "the median of the data is 35.8, and the 80% IQI of 41.2)
    - Note: the standard IQI is the middle 50% of the data., called the "Middle Quartile Range"
  
In R, you can paste together different text and numbers into a single string with the `paste()` function,
making it useful for quickly writing report text.  For example:
```{r}
paste("The mean of my data is", round(mean(bacteria_count), 1))
```


Let's try it out!

### Q
Using the quantile(), diff(), and paste() functions, Report the Median and 95% inter-quantile-range of the data.
```{r}

```


### R
Using the median, IQR() and paste() functions, report the Median and 50% IQR of the data.
```{r}

```


### S
Using the quantile() and paste() functions, Report the Middle 80th Percentile of the data.
```{r}

```

### T
Using the quantile() function, report the 0th, 25th, 50th, 75th, and 100th percentiles of the data.
```{r}

```


## Summarizing Data Visually with Percentiles: the Box Plot

Box plots are a great, simple way to represent IQRs.  In a standard box plot:
  - A line is drawn at the median (50th percentile)
  - A light box is drawn with the edges extending from the 25th percentile to the 75th percentile (the IQR)
  - "Whiskers" are drawn out from the box to the 0th and 100th percentiles...
    - ...except when the range of the data is ridculous and the whiskers are huge.  In that case, usually people will limit
      the length of the whiskers to 1.5 times the distance from the boxes, and (sometimes) just draw the data points that fall outside that range.
  - Sometimes, people will also draw the actual data points as well on top of it, for completeness.
Let's make some box plots using the `boxplot()` function!

### U

Draw a default boxplot for the bacteria count data.
```{r}

```


### V

Draw a horizontal boxplot for the bacteria count data.
```{r}

```


### X

Make the whiskers only extend out to 0.6 times the IQR.  Do "outlier points" appear?
```{r}

```


### Y
Using the `points()` function, draw the data points on top of the data.  For extra credit,
jitter the data points along the non-informative axis
so they don't overlap each other using the `jitter()` function.
```{r}

```


### Z
Here are five data sets. Use a boxplot to show all five datasets and summarize their
ranges.  Combine them together with `list()` to make it all happen in one function call.
```{r}
january <- c(5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14)
february <- c(15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24)
march <- c(25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34)
april <- c(20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29)
may <- c(35, 36, 37, 38, 39, 40, 41, 42, 43, 44, 35, 36, 37, 38, 39, 40, 41, 42, 43, 44, 35, 36, 37, 38, 39, 40, 41, 42, 43, 44)
```





## Reference
```{r}
boxplot(bacteria_count, horizontal = T)
points(bacteria_count, jitter(rep(1, length(bacteria_count)), 5), )
```


