---
title: "Categorical Descriptive Analysis: Frequency Tables"
output: html_notebook
---


# Single-Variate Categorical Analysis: Relative Proportions


A "statistic" is a summary of a dataset, almost always in number form.  
Categorical (a.k.a. "Nominal") data contains only labels, no numbers!
So, how can we create numbers from nominal data?

The most basic approach is to make **counts** of each value in the dataset. 
This is called a **"frequency statistic"**, because it tells you how frequently
a value appears in your dataset.

$$  
\tilde{c}_{Heads} = \sum\limits_{i=1}^{N} X_i 
\begin{cases} 
1 & \text{if Xi = "Heads"} \\
0 & \text{otherwise}
\end{cases}
$$
  

From the counts, you can then calculate **proportions**, sometimes called a 
**"normalized frequency"**.  This makes datasets of different sizes easier to
compare to each other.


$$  
\tilde{p}_{Heads} : \text{"The proportion of Heads in a set of coin flips"} \\
\tilde{p}_{Heads} = \sum\limits_{i=1}^{N} \frac{X_{i}}{N} = \frac{\tilde{c}_{Heads}}{N}
$$

## Calculating Proportions by first transforming the data.


Examples:
 - total number of A's: `sum(data == 'A')`
 - proportion of A's:  `mean(data == 'A')`


### Coin Flips

Given the sequence of coin flips below (Heads and Tails), 
calculate the descriptive statistics requested:
```{r}
coinFlips <- strsplit("HTHHTHHHTHTTTHTTHTHHHTHTTTHHTHTHHHTHT", "")[[1]]
coinFlips
```

How many total coin flips were there?
```{r}
length(coinFlips)
```

What were the *frequency* of "Heads" observations (i.e. how many total heads were observed), and "Tails" observations?
```{r}
length(coinFlips[coinFlips == 'H'])
sum(coinFlips == 'H')
```

What were the *proportions* (between 0 and 1) of "Heads" and "Tails" observed?
```{r}
mean(coinFlips == 'H')
```

What *percent* (between  0 and 100) of coin flips were Heads?
```{r}
mean(coinFlips == 'H') * 100
```

Make a `barplot()` showing the frequencies of Heads and Tails observations:
```{r}
heads = sum(coinFlips == 'H')
tails = sum(coinFlips == 'T')
barplot(c(Heads=heads, Tails=tails))
```


### Working with `data.frame()`: Counting cars

R comes with a built-in dataset called `mtcars`, which has
some details for different types of autos.  Instead of a vector or table, it
is stored as a `data.frame()`, which has rows of observations and columns of variables
associated with each observation. 

`data.frame` reference:
  - `dd < data.frame(heights=c(2.1, 1.7), weights=c(93, 78))`: Creating a dataframe from scratch
  - `dd['heights']`: Make a smaller dataframe with only the one column.
  - `dd[['heights']]`: Get the column 'heights' as a vector.
  - `dd$heights`: Get the column 'heights' as a vector (same as above, yes)
  - `dd[1:5, 'heights']`: Get the first 5 rows and the column 'heights' as a vector
  - `dd[1:5,]`: Get the first 5 rows
  - `dd[1:5]`: Get the first 5 columns (yes, it looks very similar.  Note the missing comma.)
  
  

Use the `mtcars` dataset (it comes with R, just use the variable name to see the data) to answer the questions below.
```{r}
nrow(mtcars)
class(mtcars)
```



What proportion of the cars in this dataset have four gears?
```{r}
sum(mtcars$gear == 4)
nrow(mtcars[mtcars$gear == 4,])
```


How many of the cars have three gears?
```{r}

```

How many of the cars have more than six cylinders?
```{r}

```


What proportion of the cars have more than 150 horsepower?
```{r}

```


What proportion of the cars have three gears?  Five?
```{r}

```



### Working with `table()`: Counting Weather Observations

This coding approach is a bit inefficient for data analysis; The more categories we have, the more code we have to write!
Instead, let's use R to create a new *object* that is designed to calculate and hold frequency table data:  `table`.
In statistics, this is called a *"Contingency Table"*.

Basic Reference for working with tables:
  - `tab <- table(x)`: Make a table from the data
  - `prop.table(tab)`: Translate the table into proportions
  - `table(x) / sum(x)`: Translate the table into proportions  (same as above)
  - `names(tab)`: Get the category names
  - `dimnames(tab)`: Get the name of the variable the category is organized in.
  - `tab[['category']]`: Subset the data from the category you want.
  - `plot(tab)`: Plot the data.
  - `barplot(tab)`: Make a barplot of the data.
  

Let's work with observations of the weather on different days, this time making a `table()`!
```{r}
weather <- c("Sunny", "Rain", "Sunny", "Clouds", "Sunny", "Clouds", "Clouds", "Sunny", "Sunny", "Sunny", "Sunny", "Sunny",  "Rain", "Sunny",  "Rain", "Clouds", "Clouds", "Clouds", "Clouds", "Sunny", "Sunny", "Sunny", "Clouds", "Sunny",  "Rain", "Sunny")
```


How many days were observed of each weather type?
```{r}
tab <- table(weather)
tab
```

Make a barplot of the observation counts:
```{r}
barplot(tab)
```


What proportion of observations were made of each weather type?
```{r}
prop.table(tab)
```


Indexing the frequency table, subset the value: how many  of days were sunny?
```{r}
tab[['Sunny']]
```




### Using `table()` with the the matcars dataset

```{r}
head(mtcars)
```

How many of each number of cylinders do the cars have?
```{r}
table(mtcars$cyl)
```


What proportion of the cars have each number of gears?
```{r}

```


Make a bar plot showing the proportion of number of gears on
the cars in this dataset.  Label the x and y axes, and make a title for the plot.  
```{r}

```


#### Iris dataset

The iris dataset contains collection of flowers of different
species.  


```{r}
head(iris)
```


How many flowers of each species are there in this dataset?
Make a bar plot, labelling everything relavant.
```{r}

```


What proportion of the flowers were of the species setosa?
(Subset practice)
```{r}

```


What proportion of the flowers were of the species virginica?
(Subset practice)
```{r}

```


### Aside: Understanding `class` transformations in R

One useful thing to know about R is that for every class (e.g. "table", "vec", "data.frame", "list", etc), 
there are different constructor functions; some just do the bare minimum to make the class,
and others do much more work, actually running calculations and transforming the data before
returning the class.  

The functions that start with `as.`, like `as.table()`, `as.data.frame()`, and `as.list()`
tend to do the least work.  So depending on how close our data is already to the desired
format, we may have to pick a different function to create the class that we want.

If you are ever unsure about what class your data is, you can use the `class()`
function to find out.
  
**Exercises**


Using `as.table()`: How many days was the bus early, on time, or late?
```{r}
nEarly <- 10
nOntime <- 5
nLate <- 2
tab <- as.table(c(Early=nEarly, Ontime=nOntime, Late=nLate))
tab
```



Using `table()` to count the data: 
```{r}
performances <- c(rep("Early", 10), rep("OnTime", 5), rep("Late", 2))
table(performances)
```

What counts of the marbles drawn from the cup were of each color?  Make a table with the data.
```{r}
green <- 10
red <- 5
yellow <- 2
marbles <- as.table(c(Green=green, Red=red, Yellow=yellow))
marbles
```


Let's pretend that we want to give this table to a function that doesn't understand
tables, but needs it to be a `data.frame` instead.  Turn the table into a dataframe:
```{r}
as.data.frame(marbles)
```


Now there's a function that wants the table as a `vector`:
```{r}
as.vector(marbles)
```


Oh, now there's one that wants it as a `list`!
```{r}
as.list(marbles)
```

