---
title: "R Notebook"
output: html_notebook
---

# Multivariate Categorical Data Analysis: Joint and Conditional Proportions


Counting individual variables gives us some information, but it's often more interesting
when we want to look at how those variables *relate* together in our data.  For example:
  - Do more people go on holiday when the weather is sunny?  (holiday_state vs weather)
  - Are art majors more or less likely to work in office jobs than math majors? (university_major vs job)
  - Does treatment A cure more patients than treatment B? (treatment vs survived)


## Joint Counts and Proportions

If we know how any given observation relates to two or more variables, we can 
show their co-occurrence with a **Contingency Table**

For example, from this "U.K High School" dataset:

|          | Men | Women |
|----------|------|--------|
| Football | 24   | 15     |
| Rugby    | 10   | 26     |
| Other    | 20   | 5      |


**Discussion**:  
  - How many men are in this dataset?
  - How many Football players?


In many contingency tables, people will add the row and/or column sums in the *margins* (edges)
of the tables to make it easier to interpret the data:

|          | Men | Women | *Sum* |
|----------|------|--------|-----|
| Football | 24   | 15     | *39*  |
| Rugby    | 10   | 26     | *36*  |
| Other    | 20   | 5      | *25*  |
| *Sum*      | *54*   | *46*     | *100* |


## Creating a Contingency Table (a.k.a. a "Cross-Tabulation")

Let's look at `table()` again, this time to see what extra help we can get from R
when we have multiple variables we want to compare against each other.


Below, I've created two vectors that create the original dataset and put them into a dataframe for organization.
No need to modify the code, just run it to get the `student.sports` variable name.


Using the `table()` function, recreate the contingency table above using the supplied vectors.
```{r}
sports <- as.table(rbind(Football=c(Men=24, Women=15), Rugby=c(10, 26), Other=c(20, 5)))
sports
```

Transpose the table using the `t()` function.  What happens to the table?
```{r}
par(mfrow=c(1, 3))

pie(sports['Football',])
pie(sports['Rugby',])
pie(sports['Other',])
```


Using the created table, call the `addmargins()` function to show the margin sums:
```{r}

```



Create a bar plot with showing the counts.  Try setting `beside` and `legend.text` to True,
and set `col` to a list of colors (the `RColorBrewer` package is great for making nice color
palettes).

(Tip: set the legend to be where you want with `barplot(args.legend = list(x = "topright", inset = c(- 0.15, 0)))`) https://statisticsglobe.com/move-position-of-barplot-legend-in-r

```{r}
# install.packages("RColorBrewer")
library("RColorBrewer")
# display.brewer.all()  # Run this to see some named color palettes the package supplies
colors <- brewer.pal(3, "Spectral")  # Get a vector of colors from the palette name of your choice.
colors
```

Make another barplot, this time:
  1. using color to represent gender and the x-axis to repreesnt the sport.
     (Tip: transposing the table is one way to do this.).

  2. instead of using `RColorBrewer` colors, let's choose our own colors from a palette
     that looks nice to us: https://coolors.co/palettes/trending.  
     You can give these hex values as a vector of strings (e.g. `c("#219ebc", "#ffb703")`).
     

Once you've made the graph, examine it a bit.  What is different about how you understand this data?
In other words, what message does this graph send in a stronger way than the barplot in the previous section?
```{r}

```


Finally, let's use the `plot()` function to plot the table.  What is being shown here?
Adding the colors you used before to this plot (also `col =`)
```{r}

```



Finally, `plot()` the transposed version of the table.  Did the plot produce what you expected?
What messages are these two plots each emphasizing?
```{r}

```


At this point, it would be good to go back and compare the bar plots and the area plots with each other.
Which of these plots:
  1. Uses screen and paper space efficiently (i.e. less wasted space)
  2. Is readable from a distance (i.e. can be printed smaller and still be understandable, so long as the text is legible)
  3. Allows you to get a natural "feel" of the relative proportions of each group?
  4. Allows you to accurately measure the counts of each group?
  5. Do you think you'll use most often when showing categorical frequency data?
  

## Joint Probability vs Conditional Probability


  - **Joint Proportion Tables** calculate proportions so that all bins add up to 100%
  - **Conditional Proportion Tables** calculate proportions so that either the rows or columns add to 100%



**Exercises**

Use the `prop.table()` function, and show the margin sums.  Does this create a:
  - joint proportion table (all the proportions add up to 1)
  - conditional proportion table (only the columns or rows add up to 1)?
```{r}

```



Use the `margins` option in the `prop.table()` function, 
setting it to either 1, 2, "sport" or "gender".  Add the margin sums to see.  
Does this create a:
  - joint proportion table (all the proportions add up to 1)
  - conditional proportion table (only the columns or rows add up to 1)?
```{r}

```


Use R to directly compare the proportion of men in football to the proportion of men in Rugby.
Which is greater?
**Tip**: To subset a two-dimensional table in R, you need to specify the rows first, then the columns.
(e.g. `myTable['Other', 'F']`)
```{r}

```


Compare the relative proportion of Football to Rugby for Women and men.
Which gender is more involved in Football?
```{r}

```


Plot the bar graph showing the joint proportions of everyone's involvement in sports
```{r}

```


Compare the sexes: Plot the bar graph showing the proportions of everyone's involvement in sports,
conditioned on their gender
```{r}

```


Compare the sports: Plot the bar graph showing the proportions of everyone's involvement in sports,
conditioned on their sport
```{r}

```



## Mtcars dataset

Compare the gears vs cylinders: Show the contingency table, including margin sums.
```{r}
head(mtcars)
```


Compare the gears vs cylinders: Plot the bar graph showing the proportions different 
numbers of gears, conditioned on the number of cylinders in the engine.
```{r}
head(mtcars)
```


## Iris Dataset

Compare the species in the built-in `iris` dataset: Show a bar plot comparing which species tend to have more sepal lengths
greater than average.
```{r}

```

