# Noise, Sampling, Confidence Intervals, Bootstrapping, and Standard Error of the Mean

When doing an experiment, the precision and consistency of our measurements plays 
a large role in how much variance there is in our data.  Some of this variance
may naturally arise out of the stochastic nature of the universe (determining whether
that is true is way, way outside the scope of this course 😉), but
all the rest of the variance that we are not interested in we can consider to be *"noise"*.

Let's take a look at the effect of noise on our data, first if it were to come from
a deterministic universe, then for a stochastic universe.

## Case 1: The Deterministic Universe.

Scenario: You're measuring the  reaction time of a subject who is supposed to press a button
whenever they see a white dot appear on the screen.  In a perfect world, this should take 
a fixed amount of time (let's say 800 milliseconds), meaning that if you measured 
it 50 times, you would get 50 samples of 800 milliseconds.  

Let's write the code for that down below, using R's built-in `rep()` function.  Make a histogram
showing the distribution of this data.  How does it look?
```{r}

```


Another way to write this data would be to imply a distribution with zero variance.
We'll be focusing in this unit on the normal distribution, so let's generate the
same data in a different way, using the `rnorm()` function.  Make the data and 
show that it's identical to the `rep()` case:
```{r}

```

## Biased Noise

### Scenario 1: Screen Update Rate
Now, let's simulate the effects of different noise on our data.  First, we'll look
at noise that adds bias to our results, meaning that it changes our estimate.  

Often, the noise is *"additive"*, meaning that it's not really related to what we collected.
For example, the computer screen might not have a totally-stable screen brightness because
of it's update rate, the LCD's backlight frequency, etc. 
Let's say that, on average, our screen updates every 16 milliseconds, meaning that
it could add anywhere from 0 to 16 milliseconds of time to the response.  Since this
could happen unpredicatably at any time, we'll uses a `uniform` distrubtion, using
the `runif()` function.  

Please calculate the response times with this screen-update-rate noise included,
and plot the histogram.  How has it changed?  What is the mean and 
standard deviation of the data?  What is the mean and standard deviation of the noise?


### Scenario 2: Participant's Attention Levels

Sometimes our participant is paying lots of attention, trying to responsd as fast
as possible, and sometimes not so much.  When they aren't paying attention, it adds even more noise!  Let's say that their attention, on average, adds anywhere from 25 to 70 millisonds
of noise to their response time.

Please factor in this attention level noise, on top of the previous noise.  
How has it changed?  What is the mean and standard deviation of the data?
Is the SD of our data larger, the same, or smaller than the SD of the 
individual sources of noise?


# Unbiased Noise

Other times, the noise has no effect on the mean, but rather only affects the
variance of the data. This is considered *unbiased* noise, and simply means that
we need more data in order to get a good estimation of the mean.  In the case
of some studies, this kind of noise doesn't exist (for example, the reaction time
scenario above; most people can't predict the future and so can't react faster 
than the appearance of the light stimulus.)

There are a lot of sources of this kind of noise, often falling in one or more 
categories:
   - *Random measurement error*: This type of noise occurs when there is a random 
      variation in the measurement process, leading to inaccurate results.
   - *Instrumental error*: This occurs when the measuring instrument has a 
      systematic error, such as a calibration error.
   - *Human error*: This type of noise is introduced when a human makes a mistake 
      during the experimental process, such as misreading a scale or recording 
      an incorrect value.
   - *Environmental noise*: This type of noise occurs when external factors 
      such as temperature, humidity, or wind, affect the experiment and produce 
      random variations in the results.
   - *Natural variability*: This type of noise occurs when there is natural variability 
      in the system being studied, such as in the case of biological experiments 
      where individuals may respond differently to a treatment.

Let's try simulating this in the case of a microscopy measurement of a fluoescent
calcium recording.  Here, the measurement lasts for 200 milliseconds, and 20
brightness measurements of a single spot are recorded.

First, let's start with the assumption that the "true" brightness of the spot
should be recorded at a value of 3.2. Please generate the data and plot its histogram:
```{r}

```

Next, let's add noise from the neuronal cell that was being recorded. During this 
time, the amount of calcium that bound to the cell membrane protein varied a bit
up and down, adding an additional deviation of 0.3 and plot the histogram:
```{r}

```

There was a tiny amount of vibration in the room from a train passing down the street
and from the car traffic, that moved the sample slightly so that we measured from
slightly different places on the cell.  Since the brightness isn't uniform across
the cell, it added another 0.3 to the deviation:
```{r}

```

The laser that shined the light, whose recording we are getting, didn't always 
hit the right spots on our fluorescent protein, adding another 0.3 to the deviation:
```{r}

```

Our camera sensor also had some measurement noise (0.3)...
```{r}

```

Our analog recording system wasn't shielded properly (0.3)...
```{r}

```

...etc. How much noise do we have, in total?  We added 0.3 5 times, does this mean
we should have a standard deviation of 1.5?
```{r}

```


# The "Box of Chocolates" Problem with Statistical Sampling: It's random!

Another way to think about this is that the data "arrives" to us pre-noisy, not 
from "noise" but just because it is supporsed to vary, that
there is no single "true" value that we should be measuring.  
We can simulate this by simply generating a lot of data with some variance already 
included. 

Here we're going to explore this by generating a lot of data that is uniformly-
distributed; in other words, there is no "center" point that it congregates toward.
From this, we'll do our experiment by "sampling" from the universal data:
```{r}
library(withr)
universe.data <- with_seed(42, runif(500000, min = 2, max = 4))
hist(universe.data, breaks=35)
dnorm

our.experiment <- sample(universe.data, 50, replace = T)
hist(our.experiment, breaks = 12)
```

In this case, the universe always has the same data, but our experiment is always changing,
because we get different data each time.  But still, our data can be used to "represent"
the data, and the goal is to give us evidence about how the universe data (now let's
call it by it's statistical name, the *"Population Distribution"*) is distributed.

## Challenge 1
Generate an Experiment with the smallest N you can, that 80% of the time will have a
mean and standard deviation within 10% of the population's mean and standard deviation.

Useful functions:
  - `sample(Nsamps, data_to_sample_from, replace=T)`
  - `replicate(Nreps, code_to_replicate)`
  
```{r}
N <- 7
mu <- mean(universe.data)
lower.bound <- .90 * mu
upper.bound <- 1.10 * mu
mean(replicate(5000, (lower.bound < (exp.mean = mean(sample(universe.data, N, replace = T)))) & (exp.mean < upper.bound)))
```

In Challenge 1, our goal was to find an experiment with a "*Statistical Power*" of 80%.
Let's do it again, this time with a power of 50%:
```{r}
N <- 7
mu <- mean(universe.data)
lower.bound <- .90 * mu
upper.bound <- 1.10 * mu
mean(replicate(5000, (lower.bound < (exp.mean = mean(sample(universe.data, N, replace = T)))) & (exp.mean < upper.bound)))
```


And with a power of 98%?
```{r}
N <- 7
mu <- mean(universe.data)
lower.bound <- .90 * mu
upper.bound <- 1.10 * mu
mean(replicate(5000, (lower.bound < (exp.mean = mean(sample(universe.data, N, replace = T)))) & (exp.mean < upper.bound)))
```


## Challenge 2: Estimate the Population Mean from the Sample Mean

One thing that's interesting about sample data is that if you get enough of it,
it will start to take on the shape of the population distribution.  This means
that once we have a decent distribution, we can start to treat our sample distribution 
as a sort of proxy for the parent distribution.

How much data do you need before the histogram of the sample starts to look similar
to the parent distribution?  Experiment below:
```{r}
our.experiment <- sample(universe.data, 50, replace = T)


```

Let's take the data you generated above and say it's our "real" data; in other 
words, we won't collect any more data.  What can we do, then?  Well, we can sample
from our sample!  This approach, fittingly, is called "*resampling*".

Use the `replicate()` and `sample()` functions to resample from your data, 
calculating the mean for each resample (remember to set `replace=T`).  
Then plot the histogram of the distribution of means you got in each resample.
Plot a line where the population mean is.  Is your distribution of resampled means
close to the true mean?  Run it many times.  Is it always the same?
Finally, what's interesting about this distribution of means?
```{r}
c(mean(our.experiment), sd(our.experiment))
resamples <- replicate(5000, sample(our.experiment, length(our.experiment), replace = T))
boot.means <- apply(resamples, 2, mean)
boot.sds   <- apply(resamples, 2, sd)

par(mfrow=c(1, 2))
hist(boot.means, breaks = 21, xlim=c(2.5, 3.5))
hist(boot.sds, breaks = 21)

```


So, what's your estimate of the population mean?  
Find the middle 95th percentile from this resampled distribution, and report
the lowest and highest range (e.g. "We estimate the mean to be between X and Y.")
This approach is called the "*percentile method for bootstrapping the estimation of the mean.*"
In shorthand, "*Bootstrapping the confidence intervals for the mean*"
```{r}
ci.mean <- quantile(boot.means, c(.025, .975))
ci.mean
((ci.mean[2] - ci.mean[1]) / 2) / 2
  
sd(our.experiment) / sqrt(length(our.experiment))
  
```

What you just estimated are called the **95% Confidence Intervals.**


## Challenge 3: Understanding the properties of the bootstrap distribution

Let's try it again, this time using a different sample for the experiment.  Try
it with a different N (much bigger, or much smaller).  How does the bootstrapped
distribution's width change when you change N, and how does this affect your
confidence intervals?  Secondly, how does the bootstrapped distribution's
width change when you change the number of resamples?
```{r}
library(boot)

ci
```
```{r}
cihalf <- sd(our.experiment) / sqrt(length(our.experiment)) * 2
c(mean(our.experiment) - cihalf, mean(our.experiment) + cihalf)
```


## Challenge 4: Estimate the Population Standard Deviation from the Sample Standard Deviation

Do the same thing, this time bootstrapping the confidence intervals for the standard deviation:
```{r}

```




# Standard Error of the Mean

When the population distribution is normally-distributed, then our sample 
distribution (with big enough sample sizes) will also be normally-distributed.
If we know that, then we can actually estimate the standard deviation of 
the bootstrapped distribution our accuracy of getting the mean
by dividing the standard deviation by the square root of sample size!

This number is called the *"Standard Error of the Mean"*:

$$ SEM = \frac{s}{\sqrt{N}} $$ 

## Challenge 5: Calculate the SEM

What is the SEM for our experiment with sample size of 30?  What is it for
a sample size of 60?
```{r}

```


## Challenge 6: Calculate the 95% CIs from the SEM

The SEM is the standard deviation of the normally-distributed bootstrap distribution.
How should we multiply modify it in order to get the 95% CIs?  
Using the SEM for an N=50 experiment, calculate the 95% CIs.
```{r}

```


## Challenge 7: Plot the CI as an error bar

Plot the CI as an error bar of a bar plot, showing 10 replications of the
experiment with N=50:
```{r}

```


