---
title: "Inference in Statistics: The Binomial Distribution"
output: html_notebook
---


# Calculating Probabilities from Known Models: The Binomial Distribution


  - `dbinom()`: Shows the probability of any given outcome: called the probability *density*. (i.e. PDF)
  - `pbinom()`: Shows the probability of getting that outcome *or less*: the cumulative probability density (i.e. CDF)
  - `qbinom()`: The inverse of the CDF, shows the outcome given a probability.
  - `rbinom()`: Generates random data according to that distribution.


## Simulating a Binomial Experiment: the `rbinom()` function

In the previous unit, we used the `sample()` and `replicate()` functions to 
simulate binomial experiment where a total number of successes were counted.  
Here, we'll simplify the code by asking the `rbinom()` function to do the same
thing.  

Using the `rbinom()` and `quantile()` functions, provide 95% Confidence Intervals 
for each of the scenarios below.


**Exercises**

Use the `rbinom()` function to help answer the questions in the scenarios below.



#### Phone Numbers

If I call 10 random phone numbers, how many men would I expect to answer?

"Successes"



If I call 10 random phone numbers in an experiment, in what proportion of experiments 
could I expect exactly 3 men to answer?
```{r}

size <- 10
prob <- 0.5
NReps <- 100000
results <- rbinom(NReps, size = size, prob = prob)
tab <- prop.table(table(results))
# barplot(tab)
tab[['3']]
mean(results == 3)

```

If I call 10 random phone numbers in an experiment, in what proportion of experiments
would I expect at least 6 men to answer?
```{r}
mean(results >= 6)
sum(tab[7:length(tab)])
```



Plot a clean distribution of possible outcomes for this experiment.
```{r}
barplot(tab)
```


What is the expected value?  Print the expected value, as well as 95% Confidence Intervals
Tip: The `quantile()` function is helpful here
```{r}
quantile(results, probs = c(.025, .1, .5, .9, .975))
```


What was the probability of our friend reaching 8 women or more in 10 random phone calls?
```{r}
mean(results >= 8)
```



#### Coin Flips


A company that sells coins for magicians: if you flip it, 
75% of the time it will show heads.  A customer calls to complain, saying that
she flipped it 16 times and saw exactly 8 heads.  They hire you to determine
if this is a valid complaint.


If I flip this coin 16 times, in what proportion of experiments would I expect to see at exactly 8 tails?
```{r}
size <- 16
prob <- 0.25
NReps <- 100000
results <- rbinom(NReps, size = size, prob = prob)
prop.table(table(results))
mean(results == 8)
```

NullHypothesis: 0.25
Alternate Hypothesis: More than 0.25
p-value = 0.027

If I flip this coin 16 times, in what proportion of experiments would I expect to see at least 8 tails?
```{r}
mean(results >= 8)
```



If I flip this coin 20 times as an experiment, what is the range of possible outcomes for this experiment?
Plot a clean distribution of outcomes.
```{r}
barplot(prop.table(table(results)))
```


What is the expected value for the coin?  
Print the expected value, as well as 95% Confidence Intervals.
How well did the customer demonstrate the coin was not as advertised? 
```{r}
quantile(results, probs = c(.025, .5, .975))
```


Given that the coin worked as advertised, what was the probability of getting 
at least the same number of heads the customer did?
```{r}
mean(results >= 8)
```



## Calculating the Expected Value Directly: the `pbinom()` and `dbinom()` functions




**Exercises**
Using the `pbinom()` and `dbinom()` functions, answer the questions in the scenarios
below




#### Phone Numbers

A friend claims that he once called 10 random phone numbers to try a new method
for meeting people to date, and 8 of them were women!  Is this a very unusual
outcome?  Let's examine the distribution.


If I call 10 random phone numbers in an experiment, in what proportion of experiments 
could I expect exactly 3 men to answer?
```{r}
plot(0:10, dbinom(0:10, size = 10, prob = 0.5), type='l')
```

If I call 10 random phone numbers in an experiment, in what proportion of experiments
would I expect at least 6 women to answer?
```{r}
# plot(0:10, pbinom(0:10, size = 10, prob = 0.5), type='h')
pbinom(0:10, size = 10, prob = 0.5)[5]
pbinom(4, size = 10, prob = 0.5)
```



Plot a clean distribution of possible outcomes for this experiment.
```{r}

```


What is the expected value?  Print the expected value, as well as 95% Confidence Intervals
```{r}
qbinom(c(.025, .5, .975), size=10, prob = 0.5)
```


What was the probability of our friend reaching 8 women or more in 10 random phone calls?
```{r}

```


#### Coin Flips


A company that sells coins for magicians: if you flip it, 
75% of the time it will show heads.  A customer calls to complain, saying that
she flipped it 16 times and saw exactly 8 heads.  They hire you to determine
if this is a valid complaint.


If I flip this coin 16 times, in what proportion of experiments would I expect to see at exactly 8 heads?
```{r}
dbinom(8, size=16, prob = 0.75)
```


If I flip this coin 16 times, in what proportion of experiments would I expect to see at least 8 heads?
```{r}
pbinom(8, size=16, prob = 0.75)
```



If I flip this coin 20 times as an experiment, what is the range of possible outcomes for this experiment?
Plot a clean distribution of outcomes.
Using `barplot()`:
```{r}

```


Using `plot()`
Type option: 
 - "p" for points, 
 - "l" for lines, 
 - "b" for both points and lines, 
 - "c" for empty points joined by lines, 
 - "o" for overplotted points and lines, 
 - "s" and "S" for stair steps and 
 - "h" for histogram-like vertical lines
```{r}

```


What is the expected value for the customer's experiment?  
Print the expected value, as well as 95% Confidence Intervals.
How well did the customer demonstrate the coin was not as advertised? 
```{r}

```


Given that the coin worked as advertised, what was the probability of getting 
at least the same number of heads the customer did?
```{r}

```



```{r}
plot(dnorm(seq(-3, 3, by = 0.1), mean = 0, sd = 1 ))
```



## The Binomial Test

It's great that we can calculate the probabilities ourselves, but there are 
also functions available to help us with this.  In this case, the 
`binom.test()` function.  Let's use it to test our calculations:



Given that the coin worked as advertised, what was the probability of getting 
at least the same number of heads the customer did?
```{r}
results <- binom.test(8, 16, p=0.25, alternative = "greater")
results$p.value
```





```{r}
binom.test(8, 10, p=0.5, alternative = "greater")
```

