---
title: "R Notebook"
output: html_notebook
---



# Multivariate Categorical Data Analysis: Joint and Conditional Proportions


Counting individual variables gives us some information, but it's often more interesting
when we want to look at how those variables *relate* together in our data.  For example:
  - Do more people go on holiday when the weather is sunny?  (holiday_state vs weather)
  - Are art majors more or less likely to work in office jobs than math majors? (university_major vs job)
  - Does treatment A cure more patients than treatment B? (treatment vs survived)


## Joint Counts and Proportions

If we know how any given observation relates to two or more variables, we can 
show their co-occurrence with a **Contingency Table**

For example, from the "U.K High School" dataset:

|          | Male | Female |
|----------|------|--------|
| Football | 24   | 15     |
| Rugby    | 10   | 26     |
| Other    | 20   | 5      |


**Discussion**:  
  - How many Males are in this dataset?
  - How many Football players?


In many contingency tables, people will add the row and/or column sums in the *margins* (edges)
of the tables to make it easier to interpret the data:

|          | Male | Female | Sum |
|----------|------|--------|-----|
| Football | 24   | 15     | 39  |
| Rugby    | 10   | 26     | 36  |
| Other    | 20   | 5      | 25  |
| Sum      | 54   | 46     | 100 |



## Creating a Contingency Table (a.k.a. a "Cross-Tabulation")

Below, I've created two vectors that create the original data.  
Using the `table()` function, recreate the contingency table above using the
supplied vectors.  (e.g. `table(var1=var1, var2=var2`)
```{r}

# Data supplied (run to see vectors):
sex = c(
  rep("M", 24), rep("F", 15), 
  rep("M", 20), rep("F", 5), 
  rep("M", 10), rep("F", 26)
)
sex

sport = c(
  rep("Football", 24), rep("Football", 15), 
  rep("Rugby", 20),    rep("Rugby", 5), 
  rep("Other", 10),    rep("Other", 26)
)
sport

# Create the table below:

tt <- table(sport=sport, sex=sex)
tt
```


Transpose the table using the `t()` function.  What happens to the table?
```{r}
t(tt)
```


Using the created table, call the `addmargins()` function to show the margin sums:
```{r}
addmargins(tt)
```



Create a bar plot with showing the counts.  Try setting `beside` and `legend.text` to True
```{r}
barplot(tt, beside = T, legend.text = T, col = brewer.pal(3, "Set32"))
# install.packages("RColorBrewer")
# library("RColorBrewer")
# display.brewer.all()
```



## Joint Probability vs Conditional Probability


Were a greater proportion footballers men or women? How does that proportion
compare with the rest of the sports?

  - **Joint Proportion Tables** calculate proportions so that all bins add up to 100%
  - **Conditional Proportion Tables** calculate proportions so that either the rows or columns add to 100%



**Exercises**

Use the `prop.table()` function, and show the margin sums.  Does this create a:
  - joint proportion table (all the proportions add up to 1)
  - conditional proportion table (only the columns or rows add up to 1)?
```{r}
addmargins(prop.table(tt))
addmargins(prop.table(tt, margin = "sex"))
addmargins(prop.table(tt, margin = "sport"))
```


Use the `margins` option in the `prop.table()` function, 
setting it to either 1, 2, "sport" or "sex".  Add the margin sums to see.  
Does this create a:
  - joint proportion table (all the proportions add up to 1)
  - conditional proportion table (only the columns or rows add up to 1)?
```{r}

```


Compare the proportion of males in football to the proportion of males in Rugby.
Which is greater?
**Tip**: To subset a two-dimensional table in R, you need to specify the rows first, then the columns.
(e.g. `myTable['Other', 'F']`)
```{r}
tt
```

```{r}
prop.table(tt[c("Football", "Other"),], margin = "sex")
```



Compare the relative proportion of Football to Rugby for Females and Males.
Which sex is more involved in Football?
```{r}

```


Plot the bar graph showing the joint proportions of everyone's involvement in sports
```{r}

```


Compare the sexes: Plot the bar graph showing the proportions of everyone's involvement in sports,
conditioned on their sex
```{r}

```


Compare the sports: Plot the bar graph showing the proportions of everyone's involvement in sports,
conditioned on their sport
```{r}

```



## Mtcars dataset

Compare the gears vs cylinders: Show the contingency table, including margin sums.
```{r}
head(mtcars)
```


Compare the gears vs cylinders: Plot the bar graph showing the proportions different 
numbers of gears, conditioned on the number of cylinders in the engine.
```{r}
head(mtcars)
```


## Iris Dataset

Compare the species: Show a bar plot comparing which species tend to have more sepal lengths
greater than average.
```{r}
iris
```



# Extra: Adding Color Palettes with RColorBrewer

https://bookdown.org/koehlerj/qr_book/rcolorbrewer-and-barplots.html

```{r}
# install.packages("RColorBrewer")
# library("RColorBrewer")
# display.brewer.all()
barplot(tt, beside = T, legend.text = T, col = brewer.pal(3, "Set32"))
```


# Extra: Moving the Legend Somewhere else

https://statisticsglobe.com/move-position-of-barplot-legend-in-r

```{r}
barplot(tt, beside = T, legend.text = T, col = brewer.pal(3, "Set32"))

        args.legend = list(x = "topright",
                           inset = c(- 0.15, 0)))
```

