---
title: "Logicals and Functions"
subtitle: "R Syntax 2"
output: html_notebook
---

## Logical

  - **Values**:
    - True: `TRUE` or `T`
    - False: `FALSE` or `F`
  - **Operators**:
    - "Is equal to": `==`
    - "Is not equal to": `!=`
    - "Is greater than": `>`
    - "Is less than" : `<`
    - "Is greater than or equal to": `>=`
    - "Is less than or equal to":  `<=`
    - "and" (both are true):  `&`   
    - "or" (one is true):   `|`
    
    **Note**: To check if two values are equal, use the double-equal sign `==`.  To assign a value to a variable name, use the single equal sign `=` or arrow `<-`.


**Exercises**

Convert the following assertions into R code for R to evaluate for truth:


"Three is not equal to four"
```{r}
3 != 4
```

"Five is greater than six"
```{r}

```

"Six times two is less than or equal to eleven"
```{r}

```


"Five to the power of three is equal to three to the power of five"
```{r}

```


Assign "x" to "10 is equal to five plus five"
```{r}
x <- 10 == 5 + 5
x
```

Assign "var" to "nine is not equal to eight"
```{r}

```


Assign "data" to "five minus four is less than zero"
```{r}

```

Assign "data2" to "ten times eight is greater or equal to pi times ten" 
(Note: pi is already a variable)
```{r}

```

Assign "var2" to "negative five times negative two is equal to negative negative ten"
```{r}

```


Assign "a" to five.
Assign "b" to seven.
"a is greater than b"
```{r}

```


Assign "a" to "five is greater than three"
Assign "b" to "two is greater than zero"
"Both a and b are true"
```{r}

```


Assign "a" to "three is greater than two times four"
Assign "b" to "three is greater than zero"
"Both a and b are true"
```{r}

```


Assign "a" to "four is greater than two times four"
Assign "b" to "three is greater than negative five"
"Either a or b is true"
```{r}

```


## Functions

Most things you can do with R work with named functions, rather than symbol operators.  For example, there is no symbol to calculate the square root; instead, you would use the `sqrt()` function like so: `sqrt(4)`.  All functions in R have the following syntax:

`output <- functionName(input)`

Let's try out some math functions to get a feel for the syntax!

Math functions:
  - Positive/Negative functions: `abs()`, `sign()`, 
  - Rounding functions: `ceiling()`, `floor()`, `round()`
  - Power Transforms: `sqrt()`, `exp()`, `log()`, `log10()`, `log2()`, `log1p()`
  - Trigonometry functions: `sin()`, `cos()`, `tan()`, `asin()`, `acos()`, etc.
  

**Tip**: Not sure what a function does?  Get the help text by putting a ? before it:
`?abs`

**Tip2**: R not responding?  If you see a + sign at the beginning of the console,
it means R doesn't think the command is finished.  Cancel the command by clicking
on the cancel and press the "Esc" key.  When you see a > sign, R is ready for new
commands.


**Exercises**

Convert the following sentences into R code.  Each sentence may involve multiple
calculations, and there may be multiple ways of getting the same answer.


Set "x" to "The absolute value of negative four"
```{r}
x <- abs(-4)
```


Set "result" to "4.2 rounded to the nearest integer is less than 4.5"
```{r}

```


Set "data" to "4.2 rounded up is less than 4.5"
```{r}

```


Set "ninety" to "the sine of pi over two is equal to one"
```{r}

```


Set "a" to "the natural log of thirty is greater or equal to 10"
```{r}

```

Set "b" to "ten to the power of the log-base-10 of eighty is equal to eighty"
```{r}

```

Set "c" to "the square root of the sum of the square of three and the square of four
is equal to the square root of twenty-five"
```{r}

```

Set "vera" equal to "e to the power of five"  (Tip: `exp()`)
```{r}

```


## Making New Functions

Sometimes you need to do the same set of calculations over an over again.  
Rather than writing out the whole code each time, you can summarise those calculations by defining your own function and using that.  Here is how you
do that in R (note the curly braces):

```
functionName <- function(input) {
    output <- input + 3
    return(output)
}

functionName(4)
functionName(5)
```

Note: the `return()` part is optional.  R will automatically return the output
of the last calculation (just like how the last calculation is automatically
printed in the notebook).  So you could write the same thing like this:

```
functionName <- function(input) {
    output <- input + 3
    output
}

functionName(4)
functionName(5)
```

Or on a single line:


```
functionName <- function(input) {input + 3}
functionName(4)
functionName(5)
```


**Exercises**

Let's get some practice making functions!  Please make the requested functions 
that pass the tests below.  If the `stopifnot()` lines are false, you will
see an error message.  If they are true, `stopifnot()` won't return anything,
and you will see the "success" message printed.


Make the function `isThree()` that returns True if the input is three, False if not.
```{r}

isThree <- function(x) {x == 3}
stopifnot(
  isThree(3) == TRUE,
  isThree(4) == FALSE
)
print("Success!")
```


Make the function `isFour()` that returns TRUE if the input is four, False if not.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  isFour(3) == FALSE,
  isFour(4) == TRUE
)
print("Success!")
```


Make the function `isPositive()` that returns TRUE if the input is greater or equal to zero, False if not.  (in this case, we'll pretend zero is positive)

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  isPositive(3) == TRUE,
  isPositive(-2) == FALSE,
  isPositive(0) == TRUE
)
print("Success!")
```


Make the function `square()` that returns the square of the input.
```{r}

# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  square(3) == 9,
  square(4) == 16
)
print("Success!")
```

Make a function `not()` that returns False if the input is True, True if False

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  not(T) == F,
  not(F) == T
)
print("Success!")
```


### Functions with multiple inputs

Functions can have as many inputs as you want!  For example:

```
add <- function(x, y) { x + y }
add(2, 5)
add(4, 6)
```

**Exercises**


Make a function `subtract()` that subtracts the second input from the first.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  subtract(5, 3) == 2,
  subtract(2, 5) == -3,
  subtract(10, 5) == 5
)
print("Success!")
```


Make a function `isGreater()` that returns True if the first input is greater than the second, False if not.

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  isGreater(3, 4) == FALSE,
  isGreater(4, 3) == TRUE,
  isGreater(4, 4) == FALSE
)
print("Success!")
```


Make a function `sumOfSquares()` that adds together the squares of each input.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  sumOfSquares(1, 2) == 5,
  sumOfSquares(2, 3) == 13,
  sumOfSquares(3, 4) == 25
)
print("Success!")
```


Make a function `average()` that calculates the mean of the three inputs (the sum of the inputs, divided by the number of inputs)
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  average(3, 3, 3) == 3,
  average(3, 4, 5) == 4,
  average(5, 0, 10) == 5
)
print("Success!")
```



