---
title: "One-Dimensional Sequences: Named Vectors"
output: html_notebook
---

# Vectors: Positional Data

Function Reminders (with some new ones):
  - `1:5`
  - `c(1, 2, 3)`
  - `seq()`
  - `seq(by=)`
  - `seq(length.out=)`
  - `rep('Hi', 3)`
  - `rev(x)`
  - `x[3]`
  - `x[2:4]`
  - `x[2:length(x)]`
  - `x[c(1, 3)]`
  - `x[x > 2]`



**Exercises**

Make a vector of values from 10 to 20.
```{r}
c(10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20)
seq(10, 20, by=1)
seq(10, 20)
10:20
```


Make a vector of values from 10 to -10.
```{r}
10:-10
```


Put the numbers 4, 8, and 10 together
```{r}
c(4, 8, 10)
```


Make a vector of 20 ones.
```{r}
rep(c("Nick", "Yiwen"), 20)
```

Make a vector of 20 twos.
```{r}
rep(2, 20)
```



Make a vector of 4 values evenly-spaced from 0 to 10
```{r}
seq(0, 10, length.out=4)
seq(0, 10, by=2.5)
```


Make a vector of every even value from 1 to 100
```{r}
offset <- 1  # offset by one to make it even when starting from odd number
seq(1 + offset, 100, 2)

```


Make a vector of 10 values evenly-spaced from -5 to 20.
```{r}
seq(-5, 20, length.out=10)
```


Get all the values from 0 to 10, evenly spaced by 0.1
```{r}
# seq(0, 10, by=0.1)
# seq(from = 0, to = 10, by = 0.1)
# seq(0, 10, 0.1)
seq(0, 10, by=0.1)
```

```{r}
?barplot
```


Using the data below, answer the following questions
```{r}
data = c(1, 5, 78, 89, 6, 54, 43, 6, 7, 8, 5, 4, 3, 4, 5, 6, 7, 8, 8, 5, 4, 14 ,3, 4, 5, 6, 7, 8)
```


Get only the first five values
```{r}
data[1:5]
head(data, 5)

```


Get only the last five values
```{r}
data[(length(data) - 4): length(data)]
tail(data, 5)
data[-(length(data) - 5) : -1]
```


Get only the values greater than 10
```{r}
data[data > 10]
subset(data, data > 10)
```


Get only the values less than 4 or greater than 50
```{r}
data[data < 4 | data > 50]
```


Get only the values greater than 20 and less than 50
```{r}
data[data > 20 & data < 50]
```

Get only the values equal to 8.
```{r}
data[data == 8]
```


Make a barplot of the data using the `barplot()` function
```{r}
barplot(data, main = "Cool Graph", sub = "Sort of", ylab = "Numbers")
```


Make a point plot of the data using the `plot()` function.
```{r}
plot(data)
```

Make a pie plot of the data using the `pie()` function.
```{r}
barplot(table(c(1, 2, 2, 3, 3, 3, 4, 4, 4, 4, 4)))
```


Put titles and axis labels (where appropriate) for the plots above.  Look 
at the help text for the functions (`?barplot`) to see which options to use.


# Labeling Individual Data Points: Named Indexing

The vectors we've used so far all have *positional* indices; each value is ordered
numerically, appearing either in the "first", "second", or "Nth" place, and to
get them we need the number index:  `data[2]`

This is fine with random sequences, but what if we want to label each data point?
Let's look at two approaches.

## Approach 1: Keep Labels in a seperate positional vector

Example:
```{r}
people <- c("John", "Dick", "Harry")
heights <- c(152, 173, 158)
```

With this approach, if we want to write the code for getting John's height, we
have to either remember that John was the first data point, or use logical indexing
to find his data:

### Stopped here

```{r}
heights[people == "John"]
```
And if we want to plot their heights?  Then we have to tell the plot function
what the labels are.
```{r}
barplot(heights, names.arg = people)
```

What if we only want to keep John and Harry's data?
Then we have to remember to remove Dick's data from both the 
heights and the labels.
```{r}
heights1 <- heights[people %in% c("John", "Harry")]
heights1

people1 <- people[people %in% c("John", "Harry")]
people1
```


**Exercises**
Let's try it out!  Make seperate "name" and "age" vectors for the following children:
  - Cindy: Age 5
  - Anna: Age 4
  - Emma: Age 6
  - Adrian: Age 8

```{r}
names = c("Cindy", "Anna", "Emma", "Adrian")
names
ages = c(5, 4, 6, 8)
ages
```


Use R to find the values:  What is Emma's age?
```{r}
ages[names == "Emma"]
```


Use R to find the values:  What is Anna's age?
```{r}
ages[names == "Anna"]
```


Use R to find the values:  What is the `mean()` age of Cindy and Emma?
```{r}
names
mean(ages[names %in% c("Cindy", "Emma")])

```


Use R to find the values:  What is the `mean()` age of Adrian and Anna?
```{r}
mean(ages[names %in% c("Adrian", "Anna")])
```


Make a bar plot of the children's ages, including title, axis labels, and labels for each bar.
```{r}
barplot(ages, names.arg = names)
```


Remove Anna from the dataset, and make a new bar plot.
```{r}
names1 <- names[names != "Anna"]
names1
ages1 <- ages[names != "Anna"]
ages1

barplot(ages1, names.arg = names1)
```


##  Approach 2: Named Vectors

Alternatively, you can keep the data values and labels together!  Insteac of having 
positional vectors, you can give each index a name:

```{r}
heights = c(John=152, Dick=173, Harry=158)
heights
```

Subsetting works by name.  
```{r}
heights['John']
```

If you want only the number from the subset, you can use the double square brackets:
```{r}
heights[['John']]
```


Plots will automatically use the names:
```{r}
barplot(heights)
```


Finally, you can separate out the names from the values with the `names()` and `as.numeric()` functions:
```{r}
names(heights)
as.numeric(heights)
```


**Exercises**
Let's try it out!  Make a named "age" vector for the following children:
  - Cindy: Age 5
  - Anna: Age 4
  - Emma: Age 6
  - Adrian: Age 8

```{r}
ages <- c(Cindy=5, Anna=4, Emma=6, Adrian=8)
ages
```


Use R to find the values:  What is Emma's age?
```{r}
ages['Emma']
ages
```


Use R to find the values:  What is Anna's age?
```{r}
ages[['Anna']]
```


Use R to find the values:  What is the `mean()` age of Cindy and Emma?
```{r}
ages[names(ages) %in% c("Cindy", "Emma")]
ages[c("Cindy", "Emma")]
ages[c(1, 3)]
```


Use R to find the values:  What is the `mean()` age of Adrian and Anna?
```{r}

```


Make a bar plot of the children's ages, including title, axis labels, and labels for each bar.
```{r}
barplot(ages)
```


Remove Anna from the dataset, and make a new bar plot.
```{r}
pie(ages)
```
  

