---
title: "Categorical Descriptive Analysis: Frequency Tables"
output: html_notebook
---


# Single-Variate Categorical Analysis: Relative Proportions


A "statistic" is a summary of a dataset, almost always in number form.  
Categorical (a.k.a. "Nominal") data contains only labels, no numbers!
So, how can we create numbers from nominal data?

The most basic approach is to make **counts** of each value in the dataset. 
This is called a **"frequency statistic"**, because it tells you how frequently
a value appears in your dataset.

There are two methods with statistical software for counting your data:

1. Directly counting each of the values:


2. Transforming the data from "categorical" to "interval" data (i.e. changing from labels to ones and zeros), then counting the ones (i.e. summing the data).
  
Though the second method may sound much more complicated than the first, many
tools make this approach straightforward,and often you'll see code that does both,
often in the same analysis.    
$$  
\tilde{c}_{Heads} = \sum\limits_{i=1}^{N} X_i 
\begin{cases} 
1 & \text{if Xi = "Heads"} \\
0 & \text{otherwise}
\end{cases}
$$
  
  
In this notebook, we'll explore both of these approaches,
and use some functions in R that make counting data straightfward.



From the counts, you can then calculate **proportions**, sometimes called a 
**"normalized frequency"**.  This makes datasets of different sizes easier to
compare to each other.


$$  
\tilde{p}_{Heads} : \text{"The proportion of Heads in a set of coin flips"} \\
\tilde{p}_{Heads} = \sum\limits_{i=1}^{N} \frac{X_{i}}{N} = \frac{\tilde{c}_{Heads}}{N}

$$

## Calculating Proportions by first transforming the data.


Examples:
 - total number of A's: `sum(data[data == 'A'])`
 - proportion of A's:  `mean(data[data == 'A'])`


#### Coin Flips

Given the sequence of coin flips below (Heads and Tails), 
calculate the descriptive statistics requested:
```{r}
flipsText <- "HTHHTHHHTHTTTHTTHTHHHTHTTTHHTHTHHHTHT"
coinFlips <- strsplit(flipsText, "")[[1]]
coinFlips
```

How many total coin flips were there?
```{r}

```

What were the frequency of "Heads" observations (i.e. how many total heads were observed), and "Tails" observations?
```{r}

```

What were the proportions (between 0 and 1) of "Heads" and "Tails" observed?
```{r}

```

What percent (between  0 and 100%) of coin flips were Heads?
```{r}

```

#### Weather

Let's work with observations of the weather on different days!
```{r}
weather <- c("Sunny", "Rain", "Sunny", "Clouds", "Sunny", "Clouds", "Clouds", "Sunny", "Sunny", "Sunny", "Sunny", "Sunny",  "Rain", "Sunny",  "Rain", "Clouds", "Clouds", "Clouds", "Clouds", "Sunny", "Sunny", "Sunny", "Clouds", "Sunny",  "Rain", "Sunny")
```


What proportion of days were rainy?
```{r}

```


What proportion of days were sunny?
```{r}

```


#### The mtcars dataset

R comes with a built-in dataset called `mtcars`, which has
some details for different types of autos.  

To get a specific varaible in the dataset, you can subset it
either with `mtcars[['varname']]` or `mtcars$varname`.

Use the `mtcars` dataset to answer the following questions:


What proportion of the cars have four gears?
```{r}
mean(mtcars$gear == 4)
```


How many of the cars have three gears?
```{r}

```

How many of the cars have more than six cylinders?
```{r}

```


What proportion of the cars have more than 150 horsepower?
```{r}

```


What proportion of the cars have three gears?  Five?
```{r}

```


## Calculating Proportions through Direct Counting


The `table()` function automatically calculates a frequency table, showing how many total observations of all events there were in a vector.  

  - Make a Frequency Table:
    - `table(x)`: Make a table of observations
  - Make a Proportion Table (2 ways):
    - `table(x) / sum(x)`:
    - `prop.table(table(observations))`: 
  - Get just one variable from the table:
    - `table(x)[['var']]`
  - Plot a table:
    - `plot(tab)`
    - `barplot(tab, main="Title", sub="Subtitle", xlab="X Axis Label", ylab="Y Axis Label")`
    

#### Coin Flips Data

Let's try it out with the coin flips!

```{r}
flipsText <- "HTHHTHHHTHTTTHTTHTHHHTHTTTHHTHTHHHTHT"
coinFlips <- strsplit(flipsText, "")[[1]]
coinFlips
```


How many of each coin side was there in this sequence of coin flips?
```{r}

```

What was the proportion of each coin side?
```{r}

```


How many total heads are shown in this table? 
(Subset practice: get the Heads result from the table)
```{r}

```


What proportion of Tails were there?
(Subset practice: get the Tails result from the table)
```{r}

```


Make a bar plot showing the number of heads and tails.
```{r}

```


Make a bar plot showing the proportion of heads and tails.
Title the plot "Coin Flip Experiment" and label the x-axis
and y-axis.
```{r}

```



Subset the Heads values from the table using the `[]`
*Tip*: To get only the number, use two sets of square brackets `[[]]`  
```{r}

```

Subset the Tails values from the table using the `[]`
```{r}

```

#### Weather Data


```{r}
weather <- c("Sunny", "Rain", "Sunny", "Clouds", "Sunny", "Clouds", "Clouds", "Sunny", "Sunny", "Sunny", "Sunny", "Sunny",  "Rain", "Sunny",  "Rain", "Clouds", "Clouds", "Clouds", "Clouds", "Sunny", "Sunny", "Sunny", "Clouds", "Sunny",  "Rain", "Sunny")
```


How many days were there of each type of weather?
```{r}

```


What proportion of the days were there of each type of weather?
```{r}

```


#### The matcars dataset


```{r}
head(mtcars)
```

How many of each number of cylinders do the cars have?
```{r}

```


What proportion of the cars have each number of gears?
```{r}

```


Make a bar plot showing the proportion of number of gears on
the cars in this dataset.  Label the x and y axes, and make a title for the plot.  
```{r}

```


#### Iris dataset

The iris dataset contains collection of flowers of different
species.  


```{r}
head(iris)
```


How many flowers of each species are there in this dataset?
Make a bar plot, labelling everything relavant.
```{r}

```


What proportion of the flowers were of the species setosa?
(Subset practice)
```{r}

```


What proportion of the flowers were of the species virginica?
(Subset practice)
```{r}

```


### Making Tables: Starting with Processed Data


It's not always necessary to store every single observation;
sometimes, you'll just have the counts (e.g. you just know 
there are 5 boys and 3 girls, but you don't know anything else).  How do you build tables with this data?

Two options:
  1. `c(rep("Boys", 5), rep("Girls", 3))` Generate raw vectors by using the `rep()` function, then use `table()` as before.
  2.  `data <- c(boys=5, girls=3)`, then convert the vector to a table with `as.table()`.
  
  
**Exercises**


What percentage of days was the bus early,on time, or late?

Make a table, Generating the raw data using the `rep()` approach.
```{r}
nEarly <- 10
nOntime <- 5
nLate <- 2


```


Make a table, Generating the raw data using the `as.table()` approach.
```{r}
nEarly <- 10
nOntime <- 5
nLate <- 2

```


What percentage of the marbles drawn from the cup were of 
each color?

Make a table, using the `rep()` approach.
```{r}
green <- 10
red <- 5
yellow <- 2

```


Make a table, using the `as.table()` approach.
```{r}
green <- 10
red <- 5
yellow <- 2

```



Of the people questioned about their sleeping habits:
  - 5 people said they usually sleep late on Sundays
  - 20 people said they usually wake up early on Sundays.
Make a bar plot representing these results, using any approach
```{r}

```



