---
title: "R Notebook"
output: html_notebook
---



## Joint Probabilities of Independent Events: Expected Values

Often, you want to make more specific predictions: 
  - "It will rain tomorrow"  **AND** "I will receive an email"
  - "It will be a boy"  **AND** "the birthday will be in June"
  
In this case, you are saying the `all()` of your predictions will be true:

```{r}
willRain = TRUE
receiveEmail = TRUE
all(willRain, receiveEmail)
```

Or you might want to make more general predictions:
  - "It will rain tomorrow"  **OR** "I will receive email tomorrow"
  - "It will be a boy"  **OR** "the birthday will be in June"

Here, you are saying you are happy if `any()` of your predictions come true:

```{r}
willRain = TRUE
receiveEmail = FALSE
any(willRain, receiveEmail)
```
How should making multiple predictions affect our expectation of their coming about?


If the two events are truly **independent** from each other (i.e. knowing one event
doesn't tell you any information about the other event), then the calculation is
simple:

$$  
P(A, B) : \text{"the Joint Probability of A and B"} \\
P(A, B) = P(A) \times P(B)
$$ 

What about the case of calculating if `any()` of the two will occur?  This is
the same as calculating the odds that **neither** will *not* occur!

$$
P(A \cup B) : \text{"the Probability of A or B"} \\
P(not A) : \text{"the Probability of A not happening} \\

\begin{aligned}
  
  P(not A) &= 1 - P(A) \\
  P(A \cup B) &= 1 - P(not A, not B) \\
  P(A \cup B) &= 1 - P(not A) \times P(not B) \\
  P(A \cup B) &= 1 - (1- P(A)) \times (1 - P(B)) \\
  
\end{aligned}
$$

**Warning**: Sometimes the probability of sampling one event influences the probability of another.  Keep it in mind for the exercises below!

**Exercises**

Calculate the expected probabilites in the example below


You flip a coin two times.  What is the probability of both landing on tails?
```{r}

```

What is the expected probability of one landing on heads, and the other on tails?
```{r}

```

What is the expected probability of one landing on heads, and a six-sided dice roll landing on one?
```{r}

```


For a classroom of 10 boys and 4 girls, you ask the teacher to send two random
students.  What is the expected probability of both students being girls?
```{r}

```

You are playing Monopoly, where rolling "doubles" (two 6-sided dice showing the 
same number) gives you an extra turn.  What is the expected probability of rolling doubles on your next turn?
```{r}

```


What is the probability of one of the two six-sided dice showing a one?
```{r}

```


In a five-day workshop with sixty people, what is the probability of at least one of them having
a birthday during the workshop?
```{r}

```
