---
title: "R Notebook"
output: html_notebook
---



## Functions with default input values

Often, functions will have some inputs "pre-set" to a certain value.
This makes the input "optional"; you don't have to give it if you are happy
with the default.  Any inputs that don't have default values will still be 
required for the function to run.

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
add <- function(x, y=1){x + y}
add(3)  # 4
add(4)  # 5
add(3, 4) # 7
add(3, y=4)  # 7
add(x=3, y=4)  # 7
add(y=4, x=3)  # 7
```


**Exercises**


Make the function `addN()` that adds 3 to whatever number is supplied by default,
but can also add other numbers.
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
addN <- function(x, y=3){x + y}

stopifnot(
  addN(4, 3) == 7,
  addN(5) == 8,
  addN(6, y=1) == 7,
  addN(7, 8) == 15
)
print("Success!")
```



Make the function `subN()` that subtracts zero from whatever number is supplied
by default, but can also subtract other numbers.

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  subN(4, 3) == 1,
  subN(5) == 5,
  subN(5, y=1) == 4,
  subN(7, 8) == -1
)
print("Success!")
```



Most functions with defaults use them as settings that change the way the function
works.  Usually, these defaults are set to TRUE or FALSE, and you can change them
to the opposite.  To do that, we'll need to use if-else statements inside the function.

Single-line example:
```
isPositive <- function(x, incZero=F){x > 0 if incZero == F else x >= 0}
isPositive(0)  # FALSE
isPositive(0, incZero=T)  # TRUE
```

Multi-line example:
```
isPositive <- function(x, incZero=F){
  if (incZero == FALSE) {
    result <- x > 0
  } else {
    result <- x >= 0
  }
  result
}
```

Another way is with the `ifelse()` function. Here's an example:
```
isPositive <- function(x, incZero=F){ifelse(incZero == F, x > 0, x >= 0)}
```


Make the function `sub2()` that subtracts the first number minus the second 
normally, but subtracts the second number minus the first if "reverse" is True 
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  sub2(3, 2) == 1,
  sub2(2, 3) == -1,
  sub2(2, 3, reverse=T) == 1,
  sub2(1, 4, reverse=T) == 3
)
print("Success!")
```


Make the function `math()` that adds two numbers if "add" is True and subtracts
them if "add" is False

```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  math(2, 3) == 5,
  math(5, 3, add=T) == 8,
  math(4, 5, add=F) == -1,
  math(1, 1, add=F) == 0
)
```


Make the function `roundnum()` that rounds a number up if "ceil" is True and
rounds it to the nearest integer if "ceil" if False
```{r}
# Function here:

# Run the Tests below to see if the function works as requested
stopifnot(
  roundnum(5.1) == 5,
  roundnum(5.9) == 6,
  roundnum(5.1, ceil=T) == 6,
  roundnum(4.2, ceil=T) == 5,
)
```




## Extra (Optional): Making Functions with a variable number of inputs

Many functions in R have a special sumbol as an input: the "Elipsis" `...`

The ... is a special variable that holds any extra positional inputs. It can
be used to create functions that accept unlimited inputs, which the function
often makes a vector out of internally for calculation.

For example, one can make a function that accepts a "data" vector:
```{r}
average <- function(data){
  mean(data)
}

average(c(1, 2, 3))  # Making the vector when the function is called.
```

Alternatively, one can make a function that builds a "data" vector internally:
```{r}
average <- function(...){
  data <- c(...)  # Making a vector inside the function.
  mean(data)
}

average(1, 2, 3)
```

You can combine the ... symbol with positional inputs and optional inputs.
If R is not sure where to put the input, it will stick it in the ...
```{r}
showInputs <- function(x, ..., y=100){
  print(x)
  print(c(...))
  print(y)
}

showInputs(1, 2, 3, 4, 5)
```


**Exercises**

Make an `add()` function that takes a variable number of inputs to add.
```{r}
# function here

# tests here
stopifnot(
  add(1) == 1,
  add(1, 2) == 2,
  add(2, 3) == 5,
  add(1, 2, 3) == 6,
  add(1, 2, 3, 4) == 10
)
print("Success!")
```




Make a `math()` function that adds all the inputs if "add" is True, and multiplies them if not.
```{r}
# function here

# tests here
stopifnot(
  math(1, 2) == 3,
  math(1, 2, add=F) == 2,
  math(1, 2, 3, 4, add=T) == 10,
  math(1, 2, 3, 4, add=F) == 24
)
print("Success!")

```



Make an `isPositive()` function that returns a vector with a value for True for each positive number, and False for each nonpositive number.
If "incZero" is True, then 0 is considered positive.  (Tip: Don't use the `ifelse()` function.)
```{r}
# function here


# tests here
stopifnot(
  identical(isPositive(1, 2), c(T, T)),
  identical(isPositive(-1, 0, 1), c(F, F, T)),
  identical(isPositive(-1, 0, 1, incZero=T), c(F, T, T))
)
print("Success!")
```

